---
title: "Final Project: Principles and Techniques of Reproducible Science"
author: "Raimundo Espejo"
date: "2025-04-07"
output:
  html_document:
  md_document:
    variant: gfm
---

### [Link to Github](https://github.com/RaimundoEspejo/Final_Project.git)

## Introduction

Refer to README file for more information including rationale. 

## Libraries and data frame

Load necessary libraries
```{r}
library(readxl)
library(ggplot2)
library(gridExtra)
library(dplyr)
library(tidyr)
library(kableExtra)
library(broom)
library(brms)
library(lme4)
library(emmeans)
library(purrr)
```

Load the data to work with using relative path:
```{r}
Data_Frame_Allele_Frequency <- read_excel("Data_Frame_Allele_Frequency.xlsx", sheet = "Table")
```

## Trend test

Apply the Cochran-Armitage trend test for each position and substrate (Eggs/Cells) combination to check if there's a significant trend over time:
```{r}
Data_Frame_Allele_Frequency$success <- Data_Frame_Allele_Frequency$`%AF` # This assigns the values from the %AF column (which represents the proportion of the alternative allele) to a new column called success.Since %AF represents proportions, each entry in success is also a proportion (a decimal between 0 and 1) rather than an integer count.
Data_Frame_Allele_Frequency$total <- 1  # Total column creation. Since %AF is a proportion, total is set to 1 to represent it as a proportion out of 1.
```
  
### Trend test for each position and substrate 

Analyze allele frequency trends over passages, grouped by Position_Genome and Passage_Substrate:
```{r}
results_trend_test <- Data_Frame_Allele_Frequency %>%
  group_by(Position_Genome, `Passage_Substrate`, `Isolate_ID`) %>% # This groups the data by Position_Genome, Passage_Substrate, and Isolate_ID. Each unique combination is treated as an individual group for analysis.
  summarise( # Creates a summary for each group considering what is below
    test_result = list(
      prop.trend.test(success * 100, rep(100, length(success)), score = `Passage_Number`)
    )
  )
results_trend_test # This table is huge. I need to obtain the important information out of it.
```

Extract p-values and test statistics:
```{r}
results_trend_test_summary <- results_trend_test %>%
  mutate(
    p_value = sapply(test_result, function(x) x$p.value),
    statistic = sapply(test_result, function(x) x$statistic)
  )
results_trend_test_summary # Now the 2 new columns are included. Still a huge table but now I can filter the important results.
```

Filter positions where the p-value is less than or equal to 0.05:
```{r}
results_trend_test_significant <- results_trend_test_summary %>%
  filter(p_value <= 0.05)
results_trend_test_significant # Now I have just the 25 significant positions where a trend is present. 
```

### Creating a table to visualize the data

Add a column for significance:
```{r}
results_trend_test_significant$Significant_Trend <- ifelse(results_trend_test_significant$p_value <= 0.05, "Yes", "No")
```

Remove the test_result column because:
```{r}
results_trend_test_significant_clean <- results_trend_test_significant %>%
  select(-test_result)
```

Sort the data frame by Position in ascending order: 
```{r}
results_trend_test_significant_clean <- results_trend_test_significant_clean %>%
  arrange(Position_Genome)
```

Create the table with distinct row colors based on Passage Substrate:
```{r}
results_trend_test_significant_clean %>%
  kable("html", caption = "Trend Test Results for Significant Positions", align = "c") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F) %>%
  row_spec(
    which(results_trend_test_significant_clean$`Passage_Substrate` == "Cells"), 
    background = "#D9EAD3"  # Light green for Cells
  ) %>%
  row_spec(
    which(results_trend_test_significant_clean$`Passage_Substrate` == "Eggs"), 
    background = "#F9CB9C"  # Light orange for Eggs
  )
```

### Creating a plot to visualize the data

Summarize Data for Trend Plot:
```{r}
trend_plot_data <- Data_Frame_Allele_Frequency %>%
  group_by(Passage_Substrate, Passage_Number) %>%
  summarise(
    mean_AF = mean(`%AF`, na.rm = TRUE),
    sd_AF = sd(`%AF`, na.rm = TRUE),
    .groups = "drop"
  )
```

Create the Trend Plot:
```{r}
ggplot(trend_plot_data, aes(x = Passage_Number, y = mean_AF, color = Passage_Substrate, group = Passage_Substrate)) +
  geom_line(linewidth = 4) +
  labs(
    title = "Allele Frequency Trends by Passage Substrate",
    x = "Passage Number",
    y = "Mean Allele Frequency",
    color = "Passage Substrate"
  ) +
  scale_x_continuous(breaks = 1:7) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14),
    legend.position = "top"
  )
```

### Trend test by isolate

Trend Test Grouped by Isolate, Position, and Substrate:
```{r}
results_trend_test_by_Isolate <- Data_Frame_Allele_Frequency %>%
  group_by(Position_Genome, Passage_Substrate, Isolate_ID) %>%
  summarise(
    test_result = list(
      prop.trend.test(success * 100, rep(100, length(success)), score = Passage_Number)
    ),
    .groups = "drop"
  )
```

Extract key results from trend test:
```{r}
results_summary_by_Isolate <- results_trend_test_by_Isolate %>%
  mutate(
    p_value = map_dbl(test_result, ~ .x$p.value),
    statistic = map_dbl(test_result, ~ .x$statistic)
  )
```

Filter significant results:
```{r}
results_trend_test_significant_by_Isolate <- results_summary_by_Isolate %>%
  filter(p_value <= 0.05)
```

Summarize data for trend plot:
```{r}
trend_plot_data_by_Isolate <- Data_Frame_Allele_Frequency %>%
  group_by(Isolate_ID, Passage_Substrate, Passage_Number) %>%
  summarise(
    mean_AF = mean(`%AF`, na.rm = TRUE),
    sd_AF = sd(`%AF`, na.rm = TRUE),
    .groups = "drop"
  )
```

Create the trend plot:
```{r}
ggplot(trend_plot_data_by_Isolate, aes(x = Passage_Number, y = mean_AF, color = Passage_Substrate, group = Passage_Substrate)) +
  geom_line(size = 3) +
  facet_wrap(~Isolate_ID) +
  labs(
    title = "Allele Frequency Trends by Isolate and Passage Substrate",
    x = "Passage Number",
    y = "Mean Allele Frequency",
    color = "Passage Substrate"
  ) +
  scale_x_continuous(breaks = 1:7) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14),
    legend.position = "top"
  )
```

## Logistic regression model 

### Logistic regression model by substrate

Create a "success" column (the number of alternative alleles) and a "failure" column (the remaining proportion):
```{r}
Data_Frame_Allele_Frequency$success <- Data_Frame_Allele_Frequency$`%AF`  # Proportion of alternative allele
Data_Frame_Allele_Frequency$failure <- 1 - Data_Frame_Allele_Frequency$success  # Proportion of reference allele
```

Convert Passage Substrate to a factor for logistic regression:
```{r}
Data_Frame_Allele_Frequency$`Passage_Substrate` <- as.factor(Data_Frame_Allele_Frequency$`Passage_Substrate`)
```

Fitting a logistic regression model using the glm() function:
```{r}
logistic_model <- glm(cbind(success, failure) ~ `Passage_Number` * `Passage_Substrate`, 
                      data = Data_Frame_Allele_Frequency, 
                      family = quasibinomial)

summary(logistic_model) # View summary of the quasibinomial model. The summary(logistic_model) output will show me the coefficients for "Passage Number" and "Passage Substrate" as well as their interaction.
```

### Extract and format coefficients

Extract and tidy the model coefficients:
```{r}
tidy_logistic_model <- broom::tidy(logistic_model)
```

Add significance stars for interpretation:
```{r}
tidy_logistic_model <- tidy_logistic_model %>%
  mutate(Significance = case_when(
    p.value < 0.001 ~ "***",
    p.value < 0.01 ~ "**",
    p.value < 0.05 ~ "*",
    TRUE ~ ""
  ))
```

Add an interpretation column with detailed explanations:
```{r}
tidy_logistic_model <- tidy_logistic_model %>%
  mutate(Interpretation = case_when(
    term == "(Intercept)" ~ "Baseline log-odds of success (%AF) in cells at passage 0. Significant at p<0.05",
    term == "Passage_Number" ~ "Passage number has a significant positive effect. For each additional passage, the odds of success (alternative allele presence) increase by about 43%.",
    term == "Passage_SubstrateEggs" ~ "Switching from Cells to Eggs significantly increases the odds of success by approximately 62.8% (exp(0.48751)).",
    term == "Passage_Number:Passage_SubstrateEggs" ~ "The interaction term is significant. The positive effect of passage number on success is weaker in eggs compared to cells.",
    TRUE ~ "N/A"
  ))
```

Display the coefficients table with interpretation:
```{r}
tidy_logistic_model %>%
  select(term, estimate, std.error, statistic, p.value, Significance, Interpretation) %>%
  kable("html", caption = "Logistic Regression Model Coefficients with Interpretation", align = "c") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = F)
```

### Get predicted probabilities for alternative allele frequency

Visualize the predicted allele frequencies over time for each substrate:
```{r}
Data_Frame_Allele_Frequency$predicted_AF <- predict(logistic_model, type = "response")
Data_Frame_Allele_Frequency$successAF <- as.integer(round(100*Data_Frame_Allele_Frequency$Allele_Frequency, 0))
Data_Frame_Allele_Frequency$trialsAF <- 100
```

Plot observed vs predicted allele frequencies over time for each Passage Substrate:
```{r}
ggplot(Data_Frame_Allele_Frequency, aes(x = `Passage_Number`, y = success, color = `Passage_Substrate`)) +
  geom_line(aes(y = predicted_AF), size = 3) +  # Increased line thickness
  scale_x_continuous(breaks = seq(min(Data_Frame_Allele_Frequency$Passage_Number), 
                                  max(Data_Frame_Allele_Frequency$Passage_Number), by = 1)) +  # X-axis with steps of 1
  labs(title = "Observed vs Predicted Alternative Allele Frequency Over Passages",  # Add a title
       y = "Alternative Allele Frequency", 
       x = "Passage Number") +
  theme_minimal()
```

### Define a function to run logistic regression 

Fit a logistic regression model:
```{r}
get_logistic_results <- function(data) {
  glm(cbind(success, total - success) ~ `Passage_Number`, data = data, family = quasibinomial)
}
```

Filter for Cells and Eggs:
```{r}
cells_data <- Data_Frame_Allele_Frequency %>%
  filter(`Passage_Substrate` == "Cells")
eggs_data <- Data_Frame_Allele_Frequency %>%
  filter(`Passage_Substrate` == "Eggs")
```

### Perform logistic regression for cells and eggs separately 

Generate separate logistic regression models for each substrate type—"Cells" and "Eggs":
```{r}
cells_results <- get_logistic_results(cells_data)
eggs_results <- get_logistic_results(eggs_data)
```

Summary of the logistic results:
```{r}
summary(cells_results)
summary(eggs_results)
```

### Logistic regression model by isolate

Scale success and failure to integer counts:
```{r}
Data_Frame_Allele_Frequency <- Data_Frame_Allele_Frequency %>%
  mutate(success_counts = round(success * 100),  # Convert success to integer counts
         failure_counts = round(failure * 100))  # Convert failure to integer counts
```

Fit a GLMM model with Isolate_ID as a random effect:
```{r}
glmm_model_by_Isolate <- glmer(
  cbind(success_counts, failure_counts) ~ `Passage_Number` * `Passage_Substrate` + (1 | Isolate_ID),
  data = Data_Frame_Allele_Frequency,
  family = binomial(link = "logit")
)

summary(glmm_model_by_Isolate) # View the summary of the GLMM model
```

Get predicted probabilities for alternative allele frequency:
```{r}
Data_Frame_Allele_Frequency$predicted_AF_by_Isolate <- predict(glmm_model_by_Isolate, type = "response")
```

Plot observed vs predicted allele frequencies over time for each Passage Substrate:
```{r}
ggplot(Data_Frame_Allele_Frequency, aes(x = `Passage_Number`, y = success, color = `Passage_Substrate`)) +
  geom_line(aes(y = predicted_AF_by_Isolate), size = 1.5) +  # Predicted trends
  facet_wrap(~Isolate_ID) +  # Create separate panels for each isolate
  scale_x_continuous(breaks = seq(min(Data_Frame_Allele_Frequency$Passage_Number), 
                                  max(Data_Frame_Allele_Frequency$Passage_Number), by = 1)) +  # X-axis for all passages
  labs(title = "Observed vs Predicted Alternative Allele Frequency Over Passages by Isolate",
       y = "Alternative Allele Frequency",
       x = "Passage Number") +
  theme_minimal()
```

## Viral load analysis 

### Mixed effects linear model for viral load

Fit the linear mixed model:
```{r}
viral_linear_mixed_model <- lmer(Viral_Load_SQ ~ `Passage_Number` * `%AF` * `Passage_Substrate` + 
                      (1 | Isolate_ID), data = Data_Frame_Allele_Frequency)

summary(viral_linear_mixed_model) # Summary of the linear mixed model
```

Perform post hoc analysis:
```{r}
posthoc_results <- emmeans(viral_linear_mixed_model, ~ Passage_Number | `%AF` * `Passage_Substrate`)

summary(posthoc_results) # Summary of the post hoc analysis
```

### Other plot for viral load

Plot for Viral Load and Allele Frequency:
```{r}
Data_Frame_Allele_Frequency_long <- Data_Frame_Allele_Frequency %>%
  pivot_longer(
    cols = c(Viral_Load_SQ, Allele_Frequency),
    names_to = "Variable",
    values_to = "Value"
  ) %>%
  mutate(
    Value = ifelse(Variable == "Allele_Frequency", Value * 12, Value),
    Variable = recode(Variable, "Viral_Load_SQ" = "Viral Load", "Allele_Frequency" = "Allele Frequency")
  )
```

Plot with no title in the legend:
```{r}
ggplot(Data_Frame_Allele_Frequency_long, aes(x = `Passage_Number`, y = Value, color = Variable)) +
  geom_smooth(aes(group = Variable), size = 2, method = "lm", formula = y ~ splines::bs(x, 3), se = F) +
  scale_y_continuous(
    name = "Viral Load",
    sec.axis = sec_axis(~ . * (1 / 12), name = "Allele Frequency (%)")
  ) +
  facet_grid(Passage_Substrate ~ Isolate_ID) +
  labs(
    title = "Trends in Viral Load and Allele Frequency Across Passages",
    x = "Passage Number"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14),
    legend.title = element_blank(),  # Removes the legend title
    legend.position = "top"
  )
```
